---
title: "Brain Expression Analysis - Project 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages}
library(tidyverse)
library(ggplot2)
library(lme4)
library(fixest)
library(grid)
library(gridExtra)
```

# Preliminary Data Analysis

## Load Data

```{r load-data}
load("brain.rda")

cat("Available objects in workspace:\n")
ls()

cat("\nExpression data shape:", dim(expression), "\n")
cat("Array metadata shape:", dim(arraymeta), "\n")
cat("Gene metadata shape:", dim(genemeta), "\n")
```

## Explore Data Structure

```{r explore-expression}
cat("Expression data shape:", dim(expression), "\n")
head(expression[1:8, 1:10])
```

```{r explore-arraymeta}
cat("Array metadata shape:", dim(arraymeta), "\n")
arraymeta[arraymeta$patient == '1834', ]
```

## Age Distribution Analysis

```{r age-analysis}
age_array <- arraymeta$age
unique_patients <- arraymeta[!duplicated(arraymeta$patient), ]
unique_age_array <- unique_patients$age

# Create age histogram
ggplot() +
  geom_histogram(aes(x = age_array, fill = "samples"),
                 alpha = 0.7, color = "black", bins = 20) +
  geom_histogram(aes(x = unique_age_array, fill = "individual"),
                 alpha = 0.7, color = "black", bins = 20) +
  geom_vline(xintercept = 70, color = "red", linewidth = 1) +
  labs(x = "Age", y = "Count", fill = "Type") +
  scale_fill_manual(values = c("samples" = "lightblue", "individual" = "darkblue")) +
  theme_minimal() +
  annotate("text", x = 72, y = max(table(age_array)) * 0.8,
           label = "age = 70", color = "red")

ggsave("age_hist.pdf", width = 8, height = 6, dpi = 300)
```

```{r patients-under-70}
unique_patients[unique_patients$age < 70, ]
```

## Sex Distribution

```{r sex-distribution}
sex_counts <- table(arraymeta$sex)
barplot(sex_counts, main = "Sex Distribution", ylab = "Count",
        col = c("lightcoral", "lightblue"))
```

## Gene Metadata

```{r gene-metadata}
head(genemeta[, 1:10])
```

## Preprocessing

Split data by brain regions and age groups:

```{r preprocessing}
head(arraymeta)
```

### Split by Regions

```{r split-regions}
# Cerebellum (cb) - use logical indexing directly
cb_mask <- arraymeta$region.cb == 1
cb_arraymeta <- arraymeta[cb_mask, ]
cb_expression <- expression[cb_mask, ]

# Anterior cingulate (ancg)
ancg_mask <- arraymeta$region.ancg == 1
ancg_arraymeta <- arraymeta[ancg_mask, ]
ancg_expression <- expression[ancg_mask, ]

# Dorsolateral prefrontal cortex (dlpfc)
dlpfc_mask <- arraymeta$region.dlpfc == 1
dlpfc_arraymeta <- arraymeta[dlpfc_mask, ]
dlpfc_expression <- expression[dlpfc_mask, ]

cat("cb:", nrow(cb_arraymeta), "rows\n")
cat("ancg:", nrow(ancg_arraymeta), "rows\n")
cat("dlpfc:", nrow(dlpfc_arraymeta), "rows\n")
```

### Create Age-Region Splits

```{r age-region-splits}
regions <- list(
  'cb' = 'region.cb',
  'ancg' = 'region.ancg',
  'dlpfc' = 'region.dlpfc'
)

splits <- list()

for (short in names(regions)) {
  col <- regions[[short]]
  in_region <- arraymeta[[col]] == 1
  is_70plus <- arraymeta$age >= 70

  # Create logical masks for the combinations
  mask_70plus <- in_region & is_70plus
  mask_u70 <- in_region & !is_70plus

  # Use logical indexing to subset both dataframes
  splits[[paste(short, '70+', sep='_')]] <- cbind(
    arraymeta[mask_70plus, ],
    expression[mask_70plus, ]
  )
  splits[[paste(short, '70-', sep='_')]] <- cbind(
    arraymeta[mask_u70, ],
    expression[mask_u70, ]
  )

  cat(sprintf("%-6s 70+: %4d rows | 70-: %4d rows\n",
              short, sum(mask_70plus), sum(mask_u70)))
}
```

### Examine ANCG 70+ Group

```{r ancg-70plus}
head(splits[['ancg_70+']])
```

### Lab Distribution by Region and Age

```{r lab-distribution-cb}
cat("CB 70- lab distribution:\n")
cat("Davis:", sum(splits[['cb_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['cb_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['cb_70-']]$lab.michigan), "\n")

cat("\nCB 70+ lab distribution:\n")
cat("Davis:", sum(splits[['cb_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['cb_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['cb_70+']]$lab.michigan), "\n")
```

```{r lab-distribution-ancg}
cat("ANCG 70- lab distribution:\n")
cat("Davis:", sum(splits[['ancg_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['ancg_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['ancg_70-']]$lab.michigan), "\n")

cat("\nANCG 70+ lab distribution:\n")
cat("Davis:", sum(splits[['ancg_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['ancg_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['ancg_70+']]$lab.michigan), "\n")
```

```{r lab-distribution-dlpfc}
cat("DLPFC 70- lab distribution:\n")
cat("Davis:", sum(splits[['dlpfc_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['dlpfc_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['dlpfc_70-']]$lab.michigan), "\n")

cat("\nDLPFC 70+ lab distribution:\n")
cat("Davis:", sum(splits[['dlpfc_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['dlpfc_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['dlpfc_70+']]$lab.michigan), "\n")
```

# Model Fitting and Results

## Linear Regression with Fixed Effects
```{r}
# Convert region indicator variables to categorical labels
region <- arraymeta %>%
    dplyr::select(region.ancg, region.cb, region.dlpfc) %>%
    max.col %>%  # Returns column index for each row
    as.character  # Convert to character labels

# Convert lab indicator variables to categorical labels
lab <- arraymeta %>%
    dplyr::select(lab.davis, lab.irvine, lab.michigan) %>%
    max.col %>%  # Returns column index for each row
    as.character  # Convert to character labels

# Create final covariate dataset
covariates <- arraymeta %>%
    dplyr::select(patient, age, sex, arrayversion) %>%  # Select base covariates
    dplyr::mutate(lab = lab, region = region) %>%  # Add categorical lab and region
    dplyr::mutate(age.indicator = ifelse(age > 70, 1, 0))  # Binary age indicator

# Filter genes with chromosome information
is.gene <- !is.na(genemeta["chrom",])
# Subset expression data to valid genes
expression.used <- expression[ ,is.gene]
# Rename gene columns for modeling
colnames(expression.used) <- paste0("gene", 1:sum(is.gene))
# Get dependent variable names
dep.used <- unname(colnames(expression.used))
# Combine data for modeling
df <- cbind(covariates, expression.used)

# Fit fixed effects regression for all genes
mod <- fixest::feols(.[dep.used] ~ age.indicator |  sex + arrayversion + lab + region, df)
```

## Results
```{r}
# Generate compact summary with robust standard errors
result.summary <- summary(mod, "compact", se = "hetero")
# Add gene symbols to results
result.summary$lhs <- genemeta["sym", is.gene]
```

### Order by Age Indicator Significance
```{r}
# Extract coefficients and standard errors from model object directly
coefs <- coef(mod)[,"age.indicator"]
se <- se(mod, se = "hetero")[,"age.indicator"]
# Calculate t-statistics (significance measure)
t_stats <- abs(coefs / se)
# Create ordered results dataframe
ordered_results <- data.frame(
  gene_symbol = genemeta["sym", is.gene],
  coefficient = coefs,
  std_error = se,
  t_statistic = t_stats,
  stringsAsFactors = FALSE
)
# Order by t-statistic (highest significance first)
ordered_results <- ordered_results[order(t_stats, decreasing = TRUE), ]
# Display top 20 most significant genes
head(ordered_results, 20)
```

### Distribution of Age Indicator Effects
```{r}
# Histogram of coefficients
ggplot(data.frame(coef = coefs), aes(x = coef)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Age Indicator Coefficients",
       x = "Coefficient", y = "Count") +
  theme_minimal()
```

```{r}
# Histogram of t-statistics of age coefficients in each gene's regression
ggplot(data.frame(t_stat = t_stats), aes(x = t_stat)) +
  geom_histogram(bins = 50, fill = "lightcoral", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 1.96, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Age Indicator T-statistics",
       x = "Absolute T-statistic", y = "Count") +
  theme_minimal()
```

## Region-Specific Regression Analysis
```{r region-specific-analysis}
# Create region-specific datasets
regions <- c("ANCG", "CB", "DLPFC")
region_results <- list()

for(region_name in regions) {
  cat("\n=== Analyzing", region_name, "===\n")
  
  # Filter data for this region
  if(region_name == "ANCG") {
    region_mask <- covariates$region == "1"
  } else if(region_name == "CB") {
    region_mask <- covariates$region == "2"  
  } else if(region_name == "DLPFC") {
    region_mask <- covariates$region == "3"
  }
  
  # Create region-specific datasets
  region_covariates <- covariates[region_mask, ]
  region_expression <- expression.used[region_mask, ]
  
  cat("Sample size:", nrow(region_covariates), "\n")
  cat("Age distribution:", table(region_covariates$age.indicator), "\n")
  
  # Combine data for modeling
  region_df <- cbind(region_covariates, region_expression)
  
  # Fit regression for this region (without region as covariate)
  region_mod <- fixest::feols(.[dep.used] ~ age.indicator | sex + arrayversion + lab, region_df)
  
  # Extract results
  region_coefs <- coef(region_mod)[,"age.indicator"]
  region_se <- se(region_mod, se = "hetero")[,"age.indicator"]
  region_t_stats <- abs(region_coefs / region_se)
  region_pvalues <- pvalue(region_mod)[,"age.indicator"]
  
  # Calculate effect sizes for this region
  region_effect_sizes <- numeric(length(region_coefs))
  for(i in 1:length(region_coefs)) {
    older_samples <- region_covariates$age >= 70
    younger_samples <- region_covariates$age < 70
    
    if(sum(older_samples) > 0 & sum(younger_samples) > 0) {
      older_mean <- mean(region_expression[older_samples, i], na.rm = TRUE)
      younger_mean <- mean(region_expression[younger_samples, i], na.rm = TRUE)
      region_effect_sizes[i] <- older_mean - younger_mean
    } else {
      region_effect_sizes[i] <- NA
    }
  }
  
# Store results
region_results[[region_name]] <- data.frame(
  gene_symbol = genemeta["sym", is.gene],
  coefficient = region_coefs,
  std_error = region_se,
  t_statistic = region_t_stats,
  p_value = region_pvalues, 
  effect_size = region_effect_sizes,
  region = region_name,
  stringsAsFactors = FALSE
)
  
  # Print top 10 genes for this region
  top_genes_region <- region_results[[region_name]][order(region_t_stats, decreasing = TRUE), ][1:10, ]
  cat("Top 10 genes in", region_name, ":\n")
  print(top_genes_region[, c("gene_symbol", "t_statistic", "effect_size")])
}
```

## Compare Results Across Regions
```{r compare-regions}
# Combine all region results
all_region_results <- do.call(rbind, region_results)

# Find genes that are significant in multiple regions using p-value
p_threshold <- 0.05
sig_genes_by_region <- list()

for(region_name in regions) {
  region_data <- region_results[[region_name]]
  sig_genes_by_region[[region_name]] <- region_data$gene_symbol[region_data$p_value < p_threshold & !is.na(region_data$p_value)]
  cat(region_name, "- Significant genes (p < 0.05):", length(sig_genes_by_region[[region_name]]), "\n")
}

# Find overlapping significant genes
common_genes <- Reduce(intersect, sig_genes_by_region)
cat("\nGenes significant in ALL regions:", length(common_genes), "\n")
if(length(common_genes) > 0) {
  print(head(common_genes, 20))
}

# Find region-specific genes (significant in only one region)
region_specific_genes <- list()
for(region_name in regions) {
  other_regions <- regions[regions != region_name]
  other_sig_genes <- unique(unlist(sig_genes_by_region[other_regions]))
  region_specific <- setdiff(sig_genes_by_region[[region_name]], other_sig_genes)
  region_specific_genes[[region_name]] <- region_specific
  
  cat("\n", region_name, "- specific genes:", length(region_specific), "\n")
  if(length(region_specific) > 0) {
    print(head(region_specific, 10))
  }
}
```

## Scatter Plot 1: Top 4 Genes from Each Region (4x3 grid)
```{r top-genes-by-region-scatter}
# Get top 4 genes from each region
regional_top_genes <- list()
for(region_name in regions) {
  region_data <- region_results[[region_name]]
  top_4 <- head(region_data[order(region_data$p_value), ], 4)
  regional_top_genes[[region_name]] <- top_4
}

# Create plots organized by region (columns)
regional_plots <- list()

# Create plots in column-wise order (region by region)
for(col in 1:3) {  # 3 regions (columns)
  region_name <- regions[col]
  top_genes <- regional_top_genes[[region_name]]
  
  for(row in 1:4) {  # 4 genes per region (rows)
    gene_name <- top_genes$gene_symbol[row]
    p_value <- top_genes$p_value[row]
    effect_size <- top_genes$effect_size[row]
    
    # Find gene index in original expression data
    gene_symbols <- genemeta["sym", is.gene]
    gene_idx <- which(gene_symbols == gene_name)
    
    # Check if gene was found
    if(length(gene_idx) == 0) {
      cat("Warning: Gene", gene_name, "not found in expression data\n")
      next
    }
    
    # Take the first match if multiple matches found
    if(length(gene_idx) > 1) {
      gene_idx <- gene_idx[1]
    }
    
    # Make sure we have the right column
    if(gene_idx > ncol(expression.used)) {
      cat("Warning: Gene index", gene_idx, "exceeds number of columns\n")
      next
    }
    
    plot_data <- data.frame(
      Expression = expression.used[, gene_idx],
      Region = case_when(
        covariates$region == "1" ~ "ANCG",
        covariates$region == "2" ~ "CB",
        covariates$region == "3" ~ "DLPFC"
      ),
      Age_Group = ifelse(covariates$age >= 70, "70+", "<70")
    )
    
    # Check if plot_data was created successfully
    if(nrow(plot_data) == 0 || any(is.na(plot_data$Expression))) {
      cat("Warning: No valid data available for gene", gene_name, "\n")
      next
    }
    
    # Create title based on ranking
    rank_text <- c("1st", "2nd", "3rd", "4th")[row]
    
    p <- ggplot(plot_data, aes(x = Region, y = Expression, fill = Age_Group)) +
      geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
      geom_point(aes(color = Age_Group), position = position_jitterdodge(jitter.width = 0.2), 
                 alpha = 0.6, size = 1.0) +
      scale_fill_manual(values = c("70+" = "#e74c3c", "<70" = "#3498db")) +
      scale_color_manual(values = c("70+" = "#c0392b", "<70" = "#2980b9")) +
      labs(title = paste0(rank_text, " in ", region_name, ": ", gene_name),
           subtitle = paste0("p = ", format(p_value, scientific = TRUE, digits = 2)),
           x = if(row == 4) "Brain Region" else "",  # Only show x-label on bottom row
           y = if(col == 1) "Expression Level" else "") +  # Only show y-label on left column
      theme_minimal() +
      theme(
        plot.title = element_text(size = 10, hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(size = 8, hjust = 0.5, color = "gray60"),
        legend.position = "none",
        axis.text.x = element_text(size = 8, angle = if(row < 4) 0 else 0),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 9),
        axis.title.y = element_text(size = 9),
        panel.grid.minor = element_blank()
      )
    
    # Calculate plot position in the list (column-wise filling)
    plot_position <- (col - 1) * 4 + row
    regional_plots[[plot_position]] <- p
  }
}

# Add column headers using textGrob
library(grid)
library(gridExtra)

# Check if we have plots to arrange
if(length(regional_plots) >= 12) {
  # Arrange in 4x3 grid with headers
  combined_regional <- grid.arrange(
    # Row 1
    regional_plots[[1]], regional_plots[[5]], regional_plots[[9]],
    # Row 2  
    regional_plots[[2]], regional_plots[[6]], regional_plots[[10]],
    # Row 3
    regional_plots[[3]], regional_plots[[7]], regional_plots[[11]],
    # Row 4
    regional_plots[[4]], regional_plots[[8]], regional_plots[[12]],
    ncol = 3, nrow = 4,
    heights = c(1, 1, 1, 1)  # Header row is smaller
  )
  
  # Save the combined plot
  ggsave("top_4_genes_by_region_4x3.pdf", combined_regional, width = 18, height = 16, dpi = 300)
  cat("Successfully created 4x3 grid with top 4 genes per region\n")
} else {
  cat("Not enough plots created - check gene names and data\n")
  cat("Created", length(regional_plots), "plots\n")
}

# Print summary
cat("\n=== 4x3 Plot Summary ===\n")
for(region_name in regions) {
  cat("\n", region_name, " top 4 genes:\n")
  top_genes <- regional_top_genes[[region_name]]
  for(i in 1:4) {
    cat(sprintf("  %d. %s (p = %.2e)\n", i, top_genes$gene_symbol[i], top_genes$p_value[i]))
  }
}
```


## Scatter Plot 2: Age vs Expression for Most Significant Genes per Region
```{r age-expression-scatter}
# Create age vs expression scatter plots for top gene from each region
age_plots <- list()

for(i in 1:length(regions)) {
  region_name <- regions[i]
  region_data <- region_results[[region_name]]
  
  # Get the most significant gene from this region
  top_gene <- region_data[order(region_data$p_value), ][1, ]
  gene_name <- top_gene$gene_symbol
  
  # Find gene index in original expression data
  gene_symbols <- genemeta["sym", is.gene]
  gene_idx <- which(gene_symbols == gene_name)
  
  # Check if gene was found
  if(length(gene_idx) == 0) {
    cat("Warning: Gene", gene_name, "not found in expression data\n")
    next
  }
  
  # Take the first match if multiple matches found
  if(length(gene_idx) > 1) {
    gene_idx <- gene_idx[1]
  }
  
  # Create region-specific data
  region_mask <- if(region_name == "ANCG") {
    covariates$region == "1"
  } else if(region_name == "CB") {
    covariates$region == "2"
  } else {
    covariates$region == "3"
  }
  
  plot_data <- data.frame(
    Age = covariates$age[region_mask],
    Expression = expression.used[region_mask, gene_idx],
    Age_Group = ifelse(covariates$age[region_mask] >= 70, "70+", "<70")
  )
  
  # Remove any rows with missing data
  plot_data <- plot_data[complete.cases(plot_data), ]
  
  # Check if we have enough data
  if(nrow(plot_data) < 5) {
    cat("Warning: Not enough data for gene", gene_name, "in region", region_name, "\n")
    next
  }
  
  # Calculate correlation
  cor_value <- cor(plot_data$Age, plot_data$Expression, use = "complete.obs")
  
  # Calculate linear model for additional stats
  lm_model <- lm(Expression ~ Age, data = plot_data)
  r_squared <- summary(lm_model)$r.squared
  p_val_cor <- cor.test(plot_data$Age, plot_data$Expression)$p.value
  
  # Set region-specific colors
  region_colors <- list(
    "ANCG" = c("70+" = "#e67e22", "<70" = "#f39c12"),
    "CB" = c("70+" = "#229954", "<70" = "#27ae60"), 
    "DLPFC" = c("70+" = "#8e44ad", "<70" = "#9b59b6")
  )
  
  p <- ggplot(plot_data, aes(x = Age, y = Expression)) +
    geom_point(aes(color = Age_Group), size = 2.5, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid", size = 1.2, alpha = 0.8) +
    geom_vline(xintercept = 70, color = "red", linetype = "dashed", alpha = 0.7, size = 1) +
    scale_color_manual(values = region_colors[[region_name]]) +
    labs(title = paste0(region_name, ": ", gene_name),
         subtitle = paste0("p = ", format(top_gene$p_value, scientific = TRUE, digits = 3), 
                          " | r = ", round(cor_value, 3),
                          " | R² = ", round(r_squared, 3)),
         x = "Age (years)", 
         y = "Expression Level",
         color = "Age Group") +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray50"),
      legend.position = "bottom",
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "gray80", fill = NA, size = 0.5)
    ) +
    guides(color = guide_legend(override.aes = list(size = 3)))
  
  # Add sample size annotation
  n_samples <- nrow(plot_data)
  n_70_plus <- sum(plot_data$Age_Group == "70+")
  n_70_minus <- sum(plot_data$Age_Group == "<70")
  
  p <- p + annotate("text", 
                   x = min(plot_data$Age) + 0.02 * (max(plot_data$Age) - min(plot_data$Age)), 
                   y = max(plot_data$Expression) - 0.05 * (max(plot_data$Expression) - min(plot_data$Expression)),
                   label = paste0("n = ", n_samples, " (", n_70_minus, " < 70, ", n_70_plus, " ≥ 70)"),
                   hjust = 0, vjust = 1, size = 3.5, color = "gray40")
  
  age_plots[[i]] <- p
}

# Check if we have plots to arrange
if(length(age_plots) > 0) {
  # Arrange age plots in a single row
  if(length(age_plots) == 3) {
    combined_age <- grid.arrange(grobs = age_plots, ncol = 3)
  } else {
    combined_age <- do.call(grid.arrange, c(age_plots, ncol = length(age_plots)))
  }
  
  # Save the age comparison plot
  ggsave("age_expression_top_genes.pdf", combined_age, width = 18, height = 6, dpi = 300)
  cat("Successfully created age vs expression plots for", length(age_plots), "regions\n")
} else {
  cat("No age plots were created\n")
}

# Print detailed summary
cat("\n=== Age vs Expression Analysis Summary ===\n")
for(i in 1:length(regions)) {
  if(i <= length(age_plots)) {
    region_name <- regions[i]
    region_data <- region_results[[region_name]]
    top_gene <- region_data[order(region_data$p_value), ][1, ]
    
    # Recalculate stats for summary
    region_mask <- if(region_name == "ANCG") {
      covariates$region == "1"
    } else if(region_name == "CB") {
      covariates$region == "2" 
    } else {
      covariates$region == "3"
    }
    
    gene_symbols <- genemeta["sym", is.gene]
    gene_idx <- which(gene_symbols == top_gene$gene_symbol)[1]
    
    if(length(gene_idx) > 0) {
      plot_data <- data.frame(
        Age = covariates$age[region_mask],
        Expression = expression.used[region_mask, gene_idx]
      )
      plot_data <- plot_data[complete.cases(plot_data), ]
      
      if(nrow(plot_data) >= 5) {
        cor_value <- cor(plot_data$Age, plot_data$Expression)
        
        cat(sprintf("\n%s:\n", region_name))
        cat(sprintf("  Gene: %s\n", top_gene$gene_symbol))
        cat(sprintf("  P-value: %.2e\n", top_gene$p_value))
        cat(sprintf("  Effect size: %.4f\n", top_gene$effect_size))
        cat(sprintf("  Age correlation: %.3f\n", cor_value))
        cat(sprintf("  Sample size: %d\n", nrow(plot_data)))
      }
    }
  }
}
```
## Region-Specific Heatmaps 
```{r region-specific-heatmaps-display}
# Function to create simple heatmap for each region
create_region_heatmap <- function(region_name, top_n_genes = 20) {
  
  # Get region-specific data
  region_mask <- if(region_name == "ANCG") {
    covariates$region == "1"
  } else if(region_name == "CB") {
    covariates$region == "2"
  } else {
    covariates$region == "3"
  }
  
  # Get top genes for this region
  region_data <- region_results[[region_name]]
  
  # Remove any duplicated genes and sort by p-value (lowest first)
  region_data <- region_data[!duplicated(region_data$gene_symbol), ]
  region_data <- region_data[order(region_data$p_value), ]
  
  # Take top genes
  top_genes_region <- head(region_data, top_n_genes)
  top_gene_names <- top_genes_region$gene_symbol
  
  cat("Selected genes for", region_name, ":", length(top_gene_names), "\n")
  
  # Find gene indices
  gene_symbols <- genemeta["sym", is.gene]
  top_gene_indices <- sapply(top_gene_names, function(gene) {
    idx <- which(gene_symbols == gene)
    if(length(idx) > 0) idx[1] else NA
  })
  
  # Remove NAs and ensure no duplicates
  valid_indices <- !is.na(top_gene_indices)
  top_gene_indices <- top_gene_indices[valid_indices]
  top_gene_names <- top_gene_names[valid_indices]
  
  # Remove any duplicate indices
  unique_indices <- !duplicated(top_gene_indices)
  top_gene_indices <- top_gene_indices[unique_indices]
  top_gene_names <- top_gene_names[unique_indices]
  
  cat("Valid unique genes for", region_name, ":", length(top_gene_names), "\n")
  
  # Extract region-specific expression data
  region_expression <- expression.used[region_mask, top_gene_indices]
  region_covariates <- covariates[region_mask, ]
  
  # Create sample info for this region
  sample_info <- data.frame(
    Age = region_covariates$age,
    Age_Group = ifelse(region_covariates$age >= 70, "70+", "<70")
  )
  
  # Sort by age
  age_order <- order(sample_info$Age)
  region_expression_sorted <- region_expression[age_order, ]
  sample_info_sorted <- sample_info[age_order, ]
  
  # Add gene names as column names
  colnames(region_expression_sorted) <- top_gene_names
  
  # Standardize gene expression (z-score), handling NAs
  expression_scaled <- scale(region_expression_sorted)
  
  # Create heatmap (display directly, no PDF)
  # Transpose for heatmap (genes as rows)
  # Reverse the order: least significant gene at top
  heatmap_t <- t(expression_scaled)
  heatmap_t <- heatmap_t[nrow(heatmap_t):1, ]  # Reverse row order
  
  # Create age group colors for column annotation
  age_group_colors <- ifelse(sample_info_sorted$Age_Group == "70+", "#e74c3c", "#3498db")
  
  # Create color palette with NA handling
  color_palette <- colorRampPalette(c("#0066cc", "#ffffff", "#cc0000"))(100)
  
  heatmap(heatmap_t,
          Colv = NA,  # No column clustering (keep age order)
          Rowv = NA,  # No row clustering (reversed order)
          scale = "none",  # Already scaled
          col = color_palette,
          na.color = "#808080",  # Gray color for NA values
          margins = c(4, 12),  # Increase left margin for gene names
          cexRow = 0.7,  # Gene name size
          cexCol = 0.1,  # Sample label size (hidden)
          main = paste0(region_name, " - Top ", length(top_gene_names), " Genes"),
          xlab = "",  # Remove x-axis label
          ylab = "Genes (by significance)",
          ColSideColors = age_group_colors,
          labCol = rep("", ncol(heatmap_t)))  # Remove column labels completely
  
  # Add legend
  legend("topright", 
         legend = c("Age < 70", "Age ≥ 70", "Low expr.", "High expr.", "NA"), 
         fill = c("#3498db", "#e74c3c", "#0066cc", "#cc0000", "#808080"),
         title = "Legend",
         cex = 0.7)
  
  # Return summary info
  return(list(
    region = region_name,
    n_samples = nrow(region_expression_sorted),
    n_genes = ncol(region_expression_sorted),
    n_young = sum(sample_info_sorted$Age < 70),
    n_old = sum(sample_info_sorted$Age >= 70),
    age_range = range(sample_info_sorted$Age),
    top_genes = top_gene_names[1:min(5, length(top_gene_names))],
    top_pvalues = top_genes_region$p_value[1:min(5, length(top_gene_names))]
  ))
}

# Create heatmaps for all three regions
cat("Creating region-specific heatmaps (displayed directly)...\n")

region_summaries <- list()
for(region_name in regions) {
  cat("\nProcessing", region_name, "...\n")
  region_summaries[[region_name]] <- create_region_heatmap(region_name, top_n_genes = 20)
}
```