---
title: "Brain Expression Analysis - Project 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages}
library(tidyverse)
library(ggplot2)
library(lme4)
library(fixest)
```

# Preliminary Data Analysis

## Load Data

```{r load-data}
load("brain.rda")

cat("Available objects in workspace:\n")
ls()

cat("\nExpression data shape:", dim(expression), "\n")
cat("Array metadata shape:", dim(arraymeta), "\n")
cat("Gene metadata shape:", dim(genemeta), "\n")
```

## Explore Data Structure

```{r explore-expression}
cat("Expression data shape:", dim(expression), "\n")
head(expression[1:8, 1:10])
```

```{r explore-arraymeta}
cat("Array metadata shape:", dim(arraymeta), "\n")
arraymeta[arraymeta$patient == '1834', ]
```

## Age Distribution Analysis

```{r age-analysis}
age_array <- arraymeta$age
unique_patients <- arraymeta[!duplicated(arraymeta$patient), ]
unique_age_array <- unique_patients$age

# Create age histogram
ggplot() +
  geom_histogram(aes(x = age_array, fill = "samples"),
                 alpha = 0.7, color = "black", bins = 20) +
  geom_histogram(aes(x = unique_age_array, fill = "individual"),
                 alpha = 0.7, color = "black", bins = 20) +
  geom_vline(xintercept = 70, color = "red", linewidth = 1) +
  labs(x = "Age", y = "Count", fill = "Type") +
  scale_fill_manual(values = c("samples" = "lightblue", "individual" = "darkblue")) +
  theme_minimal() +
  annotate("text", x = 72, y = max(table(age_array)) * 0.8,
           label = "age = 70", color = "red")

ggsave("age_hist.pdf", width = 8, height = 6, dpi = 300)
```

```{r patients-under-70}
unique_patients[unique_patients$age < 70, ]
```

## Sex Distribution

```{r sex-distribution}
sex_counts <- table(arraymeta$sex)
barplot(sex_counts, main = "Sex Distribution", ylab = "Count",
        col = c("lightcoral", "lightblue"))
```

## Gene Metadata

```{r gene-metadata}
head(genemeta[, 1:10])
```

## Preprocessing

Split data by brain regions and age groups:

```{r preprocessing}
head(arraymeta)
```

### Split by Regions

```{r split-regions}
# Cerebellum (cb) - use logical indexing directly
cb_mask <- arraymeta$region.cb == 1
cb_arraymeta <- arraymeta[cb_mask, ]
cb_expression <- expression[cb_mask, ]

# Anterior cingulate (ancg)
ancg_mask <- arraymeta$region.ancg == 1
ancg_arraymeta <- arraymeta[ancg_mask, ]
ancg_expression <- expression[ancg_mask, ]

# Dorsolateral prefrontal cortex (dlpfc)
dlpfc_mask <- arraymeta$region.dlpfc == 1
dlpfc_arraymeta <- arraymeta[dlpfc_mask, ]
dlpfc_expression <- expression[dlpfc_mask, ]

cat("cb:", nrow(cb_arraymeta), "rows\n")
cat("ancg:", nrow(ancg_arraymeta), "rows\n")
cat("dlpfc:", nrow(dlpfc_arraymeta), "rows\n")
```

### Create Age-Region Splits

```{r age-region-splits}
regions <- list(
  'cb' = 'region.cb',
  'ancg' = 'region.ancg',
  'dlpfc' = 'region.dlpfc'
)

splits <- list()

for (short in names(regions)) {
  col <- regions[[short]]
  in_region <- arraymeta[[col]] == 1
  is_70plus <- arraymeta$age >= 70

  # Create logical masks for the combinations
  mask_70plus <- in_region & is_70plus
  mask_u70 <- in_region & !is_70plus

  # Use logical indexing to subset both dataframes
  splits[[paste(short, '70+', sep='_')]] <- cbind(
    arraymeta[mask_70plus, ],
    expression[mask_70plus, ]
  )
  splits[[paste(short, '70-', sep='_')]] <- cbind(
    arraymeta[mask_u70, ],
    expression[mask_u70, ]
  )

  cat(sprintf("%-6s 70+: %4d rows | 70-: %4d rows\n",
              short, sum(mask_70plus), sum(mask_u70)))
}
```

### Examine ANCG 70+ Group

```{r ancg-70plus}
head(splits[['ancg_70+']])
```

### Lab Distribution by Region and Age

```{r lab-distribution-cb}
cat("CB 70- lab distribution:\n")
cat("Davis:", sum(splits[['cb_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['cb_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['cb_70-']]$lab.michigan), "\n")

cat("\nCB 70+ lab distribution:\n")
cat("Davis:", sum(splits[['cb_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['cb_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['cb_70+']]$lab.michigan), "\n")
```

```{r lab-distribution-ancg}
cat("ANCG 70- lab distribution:\n")
cat("Davis:", sum(splits[['ancg_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['ancg_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['ancg_70-']]$lab.michigan), "\n")

cat("\nANCG 70+ lab distribution:\n")
cat("Davis:", sum(splits[['ancg_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['ancg_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['ancg_70+']]$lab.michigan), "\n")
```

```{r lab-distribution-dlpfc}
cat("DLPFC 70- lab distribution:\n")
cat("Davis:", sum(splits[['dlpfc_70-']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['dlpfc_70-']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['dlpfc_70-']]$lab.michigan), "\n")

cat("\nDLPFC 70+ lab distribution:\n")
cat("Davis:", sum(splits[['dlpfc_70+']]$lab.davis), "\n")
cat("Irvine:", sum(splits[['dlpfc_70+']]$lab.irvine), "\n")
cat("Michigan:", sum(splits[['dlpfc_70+']]$lab.michigan), "\n")
```

# Model Fitting and Results

## Linear Regression with Fixed Effects
```{r}
# Convert region indicator variables to categorical labels
region <- arraymeta %>%
    dplyr::select(region.ancg, region.cb, region.dlpfc) %>%
    max.col %>%  # Returns column index for each row
    as.character  # Convert to character labels

# Convert lab indicator variables to categorical labels
lab <- arraymeta %>%
    dplyr::select(lab.davis, lab.irvine, lab.michigan) %>%
    max.col %>%  # Returns column index for each row
    as.character  # Convert to character labels

# Create final covariate dataset
covariates <- arraymeta %>%
    dplyr::select(patient, age, sex, arrayversion) %>%  # Select base covariates
    dplyr::mutate(lab = lab, region = region) %>%  # Add categorical lab and region
    dplyr::mutate(age.indicator = ifelse(age > 70, 1, 0))  # Binary age indicator

# Filter genes with chromosome information
is.gene <- !is.na(genemeta["chrom",])
# Subset expression data to valid genes
expression.used <- expression[ ,is.gene]
# Rename gene columns for modeling
colnames(expression.used) <- paste0("gene", 1:sum(is.gene))
# Get dependent variable names
dep.used <- unname(colnames(expression.used))
# Combine data for modeling
df <- cbind(covariates, expression.used)

# Fit fixed effects regression for all genes
mod <- fixest::feols(.[dep.used] ~ age.indicator |  sex + arrayversion + lab + region, df)
```

## Results
```{r}
# Generate compact summary with robust standard errors
result.summary <- summary(mod, "compact", se = "hetero")
# Add gene symbols to results
result.summary$lhs <- genemeta["sym", is.gene]
```

### Order by Age Indicator Significance
```{r}
# Extract coefficients and standard errors from model object directly
coefs <- coef(mod)[,"age.indicator"]
se <- se(mod, se = "hetero")[,"age.indicator"]
# Calculate t-statistics (significance measure)
t_stats <- abs(coefs / se)
# Create ordered results dataframe
ordered_results <- data.frame(
  gene_symbol = genemeta["sym", is.gene],
  coefficient = coefs,
  std_error = se,
  t_statistic = t_stats,
  stringsAsFactors = FALSE
)
# Order by t-statistic (highest significance first)
ordered_results <- ordered_results[order(t_stats, decreasing = TRUE), ]
# Display top 20 most significant genes
head(ordered_results, 20)
```

### Distribution of Age Indicator Effects
```{r}
# Histogram of coefficients
ggplot(data.frame(coef = coefs), aes(x = coef)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Age Indicator Coefficients",
       x = "Coefficient", y = "Count") +
  theme_minimal()
```

```{r}
# Histogram of t-statistics of age coefficients in each gene's regression
ggplot(data.frame(t_stat = t_stats), aes(x = t_stat)) +
  geom_histogram(bins = 50, fill = "lightcoral", color = "black", alpha = 0.7) +
  geom_vline(xintercept = 1.96, color = "red", linetype = "dashed") +
  labs(title = "Distribution of Age Indicator T-statistics",
       x = "Absolute T-statistic", y = "Count") +
  theme_minimal()
```