---
title: "Project 2 Executive Report"
author: ""
date: "2025-10-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(lme4)
library(fixest)
library(grid)
library(gridExtra)
```

```{r data preprocessing, include=FALSE}
load("brain.rda")
# Cerebellum (cb) - use logical indexing directly
cb_mask <- arraymeta$region.cb == 1
cb_arraymeta <- arraymeta[cb_mask, ]
cb_expression <- expression[cb_mask, ]

# Anterior cingulate (ancg)
ancg_mask <- arraymeta$region.ancg == 1
ancg_arraymeta <- arraymeta[ancg_mask, ]
ancg_expression <- expression[ancg_mask, ]

# Dorsolateral prefrontal cortex (dlpfc)
dlpfc_mask <- arraymeta$region.dlpfc == 1
dlpfc_arraymeta <- arraymeta[dlpfc_mask, ]
dlpfc_expression <- expression[dlpfc_mask, ]
```

# Introduction

We identified several promising genes for future study into aging in the human brain. According to our analysis, we would suggest further examiniation of 20 genes from across three sites in the human brain. From the ACG region... These genes have demonstrated statistically signficant expression between patients less than 65 and those 65 and older at the $alpha =...$ level.

These results were obtained through versions of a linear model adapted to each brain region studied in our sample, as we observed stark differences between the expression of genes based on their location in the brain. The specification, justification, and interpretation of these models are available in the following sections.

## Exploratory Data Analysis



# Methodology

# Results

# Conclusion